DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-06-21 manuscript/manuscript.tex
adding option to embed figures in ms

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man]{apa6}
\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
\date{}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
  pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
  pdflang={en-EN},
  pdfkeywords={Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\DeclareUnicodeCharacter{0301}{*************************************}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4890}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifXeTeX
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[main=english]{babel}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\fi
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis.
DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org. It features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015), data stored and available in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare the data.

@@

DateLife's core software application consists of the R package \texttt{datelife}. Its latest stable version -- v0.6.2, is available from the CRAN repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
ape (Paradis, Claude, \& Strimmer, 2004),
bold (Chamberlain et al., 2019),
geiger (Pennell et al., 2014),
paleotree (Bapst, 2012),
phyloch (Heibl, 2008),
phylocomr (Ooms \& Chamberlain, 2018),
phytools (Revell, 2012),
rotl (Michonneau, Brown, \& Winter, 2016), and
taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: creating a search query, searching a database, and summarizing results from the search.

@@

At the time of writing of this manuscript
,
DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.

@@

@@

Processing input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (Open Tree of Life Taxonomy v3.3, June 1, 2021). For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2}A, which would not have had any data otherwise.

@@

@@

\emph{\textbf{Summarizing search results.-}}
The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
By default, DateLife uses as summarizing topology the portion of OpenTree synthetic phylogeny that contains the species provided as search query.
According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
Age data from source chronograms was congruified, decreasing to 818 different age data points available (supplementray table 2) and then age summary statistics per congruified node are obtained and used as fixed secondary calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig. \ref{fig:fringillidages}).
Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.

@@

\leavevmode\vadjust pre{\hypertarget{ref-beresford2005african}{}}%
Beresford, P., Barker, F., Ryan, Pg., \& Crowe, T. (2005). African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary {`enigmas.'} \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%    TABLE 1     %%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \caption{Ages of congruified nodes belonging to step C2 from Figure3}
  \label{tbl:table1}
  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%    TABLE 2     %%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \caption{Summary of congruified nodes ages, corresponding to step C3 from Figure3}
  \label{tbl:table2}
  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
\end{table}

% in this file I leave the figure captions outside\ caption{} because I want them
% to be formatted in the same way as the general text (double spaced and linenumbered)
\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 1    %%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\includegraphics{../figures/figure1/figure1-new.pdf}
\caption{Stylized DateLife workflow. This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}. Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
}
\label{fig:workflow}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE runtime    %%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names. We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray). We also performed a search using the already processed query for comparison (light blue).}
\label{fig:runtime_main}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2    %%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{../figures/figure2/figure2.pdf}
\caption{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes. B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names. Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}
\label{fig:figure2}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-1  %%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{figure}[!h]
%\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
%\caption{Age data results of a DateLife search of a small sample of 6 %bird species within the Passeriformes. Input names were found across 9 %chronograms within 6 independent studies (Barker et al. %(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al. %(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al. %(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al. %(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price %(\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al. %(\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data %points for the queried species names.}
%\label{fig:figure2-1}
%\end{figure}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-2  %%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{figure}[!h]
%\includegraphics{../figures/figure2/figure2-2.pdf}
%\caption{Summarized age data is used as secondary calibrations to date %a tree topology as a summary chronogram.}
%\label{fig:summaries}
%\end{figure}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
  pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
  pdflang={en-EN},
  pdfkeywords={Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\DeclareUnicodeCharacter{0301}{*************************************}

% leave the figure captions outside\ caption{} because I want them
% to be formatted in the same way as the general text (double spaced and linenumbered)

\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
\usepackage{caption}
\usepackage{float}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4906}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
\date{}

@@

Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis, for research, teaching, and science communication and policy.
DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org. It features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015), data stored and available in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare the data.

@@

DateLife's core software application consists of the R package \texttt{datelife}. Its latest stable version -- v0.6.4, is available from the CRAN repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
ape (Paradis, Claude, \& Strimmer, 2004),
bold (Chamberlain et al., 2019),
geiger (Pennell et al., 2014),
paleotree (Bapst, 2012),
phyloch (Heibl, 2008),
phylocomr (Ooms \& Chamberlain, 2018),
phytools (Revell, 2012),
rotl (Michonneau, Brown, \& Winter, 2016), and
taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: creating a search query, searching a database, and summarizing results from the search.

\begin{center}
\includegraphics[width = 6.5in]{../figures/figure1/figure1-new.pdf}
\captionof{figure}{Stylized DateLife workflow. This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website at \url{http://www.datelife.org/}. Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.}
\label{fig:workflow}
\end{center}

@@

At the time of writing of this manuscript
(Jun 20, 2022),
DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.

@@

\begin{center}
\includegraphics[width= 6.5in]{../figures/fig_runtime_main.pdf}
\end{center}
\captionof{figure}{Computation time used to process a datelife query and a datelife search across \texttt{datelife}'s chronogram database, relative to number of input taxon names. For each N = {10, 100, 200, ..., 1 000, ... , 9 000, 10 000}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search with input processing, without implementing Taxon Names Resolution Service (TNRS; dark gray), and implementing TNRS (light gray). We also performed a search using a query that had been pre-processed with TNRS, for comparison (light blue).}

\label{fig:runtime_main}

@@

Processing input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (Open Tree of Life Taxonomy v3.3, June 1, 2021). For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2}A, which would not have had any data otherwise.
\vspace{5mm}

@@

\vspace{10mm}

\begin{center}
\includegraphics[width = 6.5in]{../figures/figure2/figure2.pdf}
\end{center}
\captionof{figure}{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes. B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names. Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}

\label{fig:figure2}

\vspace{15mm}

\begin{table}[H]
  \caption{Ages of congruified nodes obtained from a datelife search for the Small Example presented. See Figure 3, step C2.}
  \label{tbl:table1}
  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
\end{table}

\begin{table}[H]
  \caption{Summary of congruified nodes ages. See Figure 3, step C3.}
  \label{tbl:table2}
  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
\end{table}

@@

\vspace{5mm}

\emph{\textbf{Summarizing search results.-}}
The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
By default, DateLife uses as summarizing topology the portion of OpenTree synthetic phylogeny that contains the species provided as search query.
According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).

Age data from source chronograms was congruified, decreasing to 818 different age data points available (supplementray table 2) and then age summary statistics per congruified node are obtained and used as fixed secondary calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig. \ref{fig:fringillidages}).
Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.

\vspace{20mm}

\begin{center}
\includegraphics[width = 6.5in]{../figures/fringillidae-topologies/fringillidae-topology.pdf}
\captionof{figure}{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
B) Topology of 289 tips and 253 internal nodes, encompassing bird species within the Fringillidae only.
This tree results from pruning species that are not assigned to the family Fringillidae (purple branches in topology A) but that form a monophyletic group with them.
Bird species within the Fringillidae are paraphyletic
(Alström et al. \protect\hyperlink{ref-alstrom2014discovery}{2014},
Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
Barker et al. \protect\hyperlink{ref-barker2013going}{2013},
Barker \protect\hyperlink{ref-barker2014mitogenomic}{2014},
Barker et al. \protect\hyperlink{ref-barker2015new}{2015},
Beresford, Barker, Ryan, \& Crowe \protect\hyperlink{ref-beresford2005african}{2005},
Bryson Jr et al. \protect\hyperlink{ref-bryson2014diversification}{2014},
Burleigh, Kimball, \& Braun \protect\hyperlink{ref-burleigh2015building}{2015},
Burns et al. \protect\hyperlink{ref-burns2014phylogenetics}{2014},
Chaves, Hidalgo, \& Klicka \protect\hyperlink{ref-chaves2013biogeography}{2013},
Claramunt \& Cracraft \protect\hyperlink{ref-claramunt2015new}{2015},
Gibb et al. \protect\hyperlink{ref-gibb2015new}{2015},
Hackett et al. \protect\hyperlink{ref-hackett2008phylogenomic}{2008},
Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012},
Johansson, Fjeldså, \& Bowi \protect\hyperlink{ref-johansson2008phylogenetic}{200},
Kimball et al. \protect\hyperlink{ref-kimball2019phylogenomic}{2019},
Klicka et al. \protect\hyperlink{ref-klicka2014comprehensive}{2014},
Lamichhaney et al. \protect\hyperlink{ref-lamichhaney2015evolution}{2015},
Lerner, Meyer, James, Hofreiter, \& Fleischer \protect\hyperlink{ref-lerner2011multilocus}{2011},
Lovette et al. \protect\hyperlink{ref-lovette2010comprehensive}{2010},
Moyle et al. \protect\hyperlink{ref-moyle2016tectonic}{2016},
Ödeen, Håstad, \& Alström \protect\hyperlink{ref-odeen2011evolution}{2011},
Oliveros et al. \protect\hyperlink{ref-oliveros2019earth}{2019},
Päckert et al. \protect\hyperlink{ref-packert2012horizontal}{2012},
Parchman, Benkman, \& Mezquida \protect\hyperlink{ref-parchman2007coevolution}{2007},
Powell et al. \protect\hyperlink{ref-powell2014comprehensive}{2014},
Price et al. \protect\hyperlink{ref-price2014niche}{2014},
Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka \protect\hyperlink{ref-pulgarin2013multilocus}{2013},
Selvatti, Gonzaga, \& Moraes Russo \protect\hyperlink{ref-selvatti2015paleogene}{2015},
Tietze, Päckert, Martens, Lehmann, \& Sun \protect\hyperlink{ref-tietze2013complete}{2013},
Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).
}
\label{fig:fringillidae-topologies}
\end{center}

@@

\leavevmode\vadjust pre{\hypertarget{ref-beresford2005african}{}}%
Beresford, P., Barker, F., Ryan, Pg., \& Crowe, T. (2005). African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary {``enigmas.''} \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.

@@
